---
title: "Documentation of DILI Scoring System"
output:
  pdf_document: default
  word_document: default
  author: Haochen Li
date: "2026-01-07"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(eval = FALSE)
```

# Overview

## Background and Motivation 
Prior to this work, DILI assessment relied heavily on subjective interpretation, particularly through manual inspection of data visualizations such as histograms. This approach has several limitations:

1. High subjectivity\
Histogram construction (e.g., binning strategy) is inherently subjective and sensitive to experimental variance, which can substantially influence interpretation.

2. Limited comparability across compounds\
When toxicity differences between compounds are subtle, it becomes difficult to make clear, quantitative comparisons using individual assay readouts or visual inspection alone.

3. Over-reliance on individual concentration levels\
In conventional analyses, concentration levels that do not reach statistical significance are often excluded, unless a strong overall dose-dependent trend is observed. As a result, potentially informative signals may be overlooked.

## Objective

This code is designed to construct and evaluate a DILI scoring system based on organchip experimental data, with the goal of providing a quantitative and comparable assessment of compound-level hepatotoxic risk.

Specifically, the system takes multiple biological assay readouts as inputs—including but not limited to cell viability, albumin, ATP, and enzymatic markers—measured across different treatment conditions and concentration levels. These data are systematically integrated, normalized, and transformed to produce interpretable compound-level DILI scores.

The resulting scores can be used to:

- Rank and compare hepatotoxic risk across compounds;

- Assess toxicity differences under different treatment conditions or experimental setups;

- Evaluate consistency and reproducibility across experimental batches or laboratories.


```{r}
## ===============================
## Package installation & loading
## This step is to load all the necessary packages for future code
## ===============================

# CRAN packages
cran_packages <- c(
  "arrow", "tidyr", "purrr", "readr", "igraph", "readxl",
  "httr", "jsonlite", "webchem", "pheatmap", "tibble",
  "stringr", "parallel", "pbmcapply", "furrr", "progressr",
  "ggplot2", "dplyr", "rlang"
)

# Get installed packages once
installed_pkgs <- rownames(installed.packages())

# Install missing CRAN packages
cran_to_install <- setdiff(cran_packages, installed_pkgs)

if (length(cran_to_install) > 0) {
  install.packages(cran_to_install, dependencies = TRUE)
}

# Load all packages
invisible(
  lapply(cran_packages, library, character.only = TRUE)
)

```


```{r}
library(arrow)
library(tidyr)
library(purrr)
library(readr)
library(igraph)
library(biomaRt)
library(readxl)
library(httr)
library(jsonlite)
library(webchem)
library(pheatmap)
library(tibble)
library(clusterProfiler)
library(ReactomePA)
library(org.Hs.eg.db)
library(stringr)
library(parallel)
library(pbmcapply)
library(furrr)
library(progressr)
library(ggplot2)
library(dplyr)
```

```{r}


#Load the data(make sure all the column names match the sample data)
# Make sure this Rmarkdown file is under the same folder as the data we want to analyse
raw_data <- read_xlsx("Data_example_2.xlsx")


raw_data <- raw_data %>%
# Remove records with invalid or unusable Drug/Code values; update the values inside 
#  quotes if different naming is used
  dplyr::filter(!`Drug/Code` %in% c("OF","UF","Bubble","leak", "Pre-treatment")) %>%

#Modify column names for future code
  dplyr::rename(Drug = `Drug/Code`,
         Concentration =`Cmax Multiplier`,
         ALB_CV = `ALB CV`,
         ALT_CV = `ALT CV`,
         Viability = `Percent Viability (%)`,
         Fold_Viability = `Fold Percent Viability`,
         ALT = `ALT (ng/day/million cells)`,
         ALB = `Albumin (µg/day/million cells)`)
raw_data

# Select the column that we care about
data <- raw_data %>% 
  dplyr::select(-Type, -Condition, -Donor, -`Cell count`, -`ALB OD1`, -`ALB OD2`
                , -`ALT OD1`,-`ALT OD2` )
data



# Build Treatment column for to seperate drugs with different concentration levels
data <- data %>%
  mutate(
    Drug          = str_trim(Drug),
    Concentration = str_trim(Concentration),

    # Remove trailing "XCmax" if present
    conc_clean = if_else(
      is.na(Concentration), NA_character_,
      str_trim(str_replace(Concentration, "(?i)\\s*XCmax\\s*$", ""))
    ),

    # Remove all whitespaces in drug names
    drug_clean = str_replace_all(Drug, "\\s+", ""),

    # Construct treatment label
    treatment = case_when(
      is.na(conc_clean) | conc_clean == "" ~
        drug_clean,

      `Vehicle control` == "Negative control" ~
  str_c(
    drug_clean,
    paste0(as.numeric(conc_clean) * 100, "%"),
    sep = "_"
  ),

      TRUE ~
        str_c(drug_clean, conc_clean, sep = "_")
    )
  ) %>%
  dplyr::select(-conc_clean, -drug_clean)


#Use this code to convert ALB_CV and ALT_CV from percentage strings (e.g., "12%") 
#to numeric proportions (e.g., 0.12) when needed for downstream calculations
# data <- data %>%
#   mutate(
#     ALB_CV = as.numeric(str_remove(ALB_CV, "%")) / 100,
#     ALT_CV = as.numeric(str_remove(ALT_CV, "%")) / 100
#   )


#Convert data into numbers for future use
data <- data %>%
  mutate(
    ALB       = suppressWarnings(as.numeric(ALB)),
    Viability = suppressWarnings(as.numeric(Viability)),
    ALT = suppressWarnings(as.numeric(ALT)))
```


```{r}

#Use this code to give a general normality check for data, also see how many 
#replicate each drug has so that we can determine the method of outlier detection we'll use
normality_results <- data %>%
  filter(!is.na(ALB)#, !is.na(ALB_CV), ALB_CV <= 0.2
         , Day !=7) %>%
  group_by(treatment, Day) %>%
  summarise(
    n = n(),
    p_value = tryCatch(shapiro.test(ALB)$p.value, error = function(e) NA_real_),
    is_normal = p_value > 0.05
  )

normality_results
```


```{r}
#Outlier detection method, it'll detect the sample size and choose the 
#corresponding method for outlier
dixon_iter_rm <- function(x, alpha = 0.05, min_n = 3, mad_thresh = 3.5) {
  n <- length(x)
  removed <- rep(FALSE, n)

  if (sum(!is.na(x)) < min_n) return(removed)

  if (sum(!is.na(x)) >= 25) {
    xx  <- x
    med <- median(xx, na.rm = TRUE)
    md  <- mad(xx, na.rm = TRUE)

    if (is.na(md) || md == 0) {

      q <- quantile(xx, probs = c(0.25, 0.75), na.rm = TRUE)
      iqr <- q[2] - q[1]
      if (is.na(iqr) || iqr == 0) return(removed)
      removed <- (xx < (q[1] - 1.5 * iqr)) | (xx > (q[2] + 1.5 * iqr))
    } else {
      z <- abs(xx - med) / md
      removed <- z > mad_thresh
      removed[is.na(removed)] <- FALSE
    }
    return(removed)
  }

  repeat {
    idx <- which(!removed & !is.na(x))
    if (length(idx) < min_n) break

    xs <- x[idx]

    tst <- tryCatch(outliers::dixon.test(xs, two.sided = TRUE),
                    error = function(e) NULL)
    if (is.null(tst)) break

    if (!is.na(tst$p.value) && tst$p.value < alpha) {
      alt <- as.character(tst$alternative)
      if (grepl("lowest|min", alt, ignore.case = TRUE)) {
        remove_idx <- idx[which.min(xs)]
      } else if (grepl("highest|max", alt, ignore.case = TRUE)) {
        remove_idx <- idx[which.max(xs)]
      } else {
          break
      }
      removed[remove_idx] <- TRUE
    } else {
      break
    }
  }
  removed
}
```

```{r}
# Generate cleaned dataset for future use(Outlier removed) 
data_alb_flagged_14 <- data %>%
  filter(!is.na(ALB), Day == 14) %>% 
  #(Add more restriction here if needed, detailed example check below)
  group_by(treatment) %>%
  group_modify(~ mutate(.x, dixon_outlier = dixon_iter_rm(.x$ALB))) %>%
  ungroup()  %>% 
  #Stop here if you want to check which compound has been labeled as outlier
  filter(!dixon_outlier) 
  #This line is to remove those compounds that have been labeled as outleir



# data_alb_flagged_11 <- data %>%
#   filter(!is.na(ALB), !is.na(ALB_CV), ALB_CV <= 0.2, Day == 11) %>%
#   group_by(treatment) %>%
#   group_modify(~ mutate(.x, dixon_outlier = dixon_iter_rm(.x$ALB))) %>%
#   ungroup() %>% 
#   filter(!dixon_outlier)


data_alt_flagged <- data %>%
  filter(!is.na(ALT), Day == 9) %>%
  group_by(treatment) %>%
  group_modify(~ mutate(.x, dixon_outlier = dixon_iter_rm(.x$ALT))) %>%
  ungroup() %>% 
  filter(!dixon_outlier)

data_atp_flagged <- data %>%
  filter(!is.na(ATP), Day == 14) %>%
  group_by(treatment) %>%
  group_modify(~ mutate(.x, dixon_outlier = dixon_iter_rm(.x$ATP))) %>%
  ungroup()%>% 
  filter(!dixon_outlier)
```

```{r}
# Function to convert results into wide table
make_level_wide <- function(df, value_col, prefix) {
  val_sym <- rlang::sym(value_col)

  df %>%
    mutate(
      Compound = sub("_[^_]*$", "", treatment),

      conc_str = sub(".*_", "", treatment),

      conc_num   = suppressWarnings(as.numeric(conc_str)),
      conc_label = ifelse(is.na(conc_num), "Base", paste0(conc_str, "X"))
    ) %>%
    arrange(Compound, conc_num) %>%
    dplyr::select(Compound, conc_label, value = !!val_sym) %>%
    tidyr::pivot_wider(
      names_from   = conc_label,
      values_from  = value,
      names_prefix = paste0(prefix, "_"),
      names_sort   = FALSE
    )
}

```

```{r}
# Function to calculate scores for each treatment
#input of the function: (name of the data, Variable name)
#Examples check below
ratio_overall_by_treatment_ttest <- function(data, response_col) {
  library(dplyr)
  library(rlang)
  library(stringr)

  resp <- ensym(response_col)
  resp_name <- as_string(resp)
  data <- data %>%
    mutate(
      `Vehicle control` = str_trim(`Vehicle control`),
      Drug              = str_trim(Drug)
    )

  out <- data %>%
    group_by(treatment) %>%
    group_modify(~{
      current_treat <- .y$treatment[[1]]


      x <- .x[[resp_name]]
      x <- x[!is.na(x)]
      n_treat    <- length(x)
      mean_treat <- if (n_treat > 0) mean(x) else NA_real_

      vc_raw <- unique(str_trim(.x[["Vehicle control"]]))
      vc_raw <- vc_raw[!is.na(vc_raw)]

  
      is_control_treat <- (length(vc_raw) == 1 && vc_raw == "Negative control")


      vc_name <- if (!is_control_treat) vc_raw[vc_raw != "Negative control"][1] else NA_character_

  
      ctrl_df <- if (!is_control_treat && !is.na(vc_name)) {
        data %>%
          filter(
            `Vehicle control` == "Negative control",
            treatment == vc_name
          )
      } else {
        data[0, , drop = FALSE]  
      }

      ctrl_vals <- ctrl_df[[resp_name]]
      ctrl_vals <- ctrl_vals[!is.na(ctrl_vals)]
      n_ctrl    <- length(ctrl_vals)
      mean_ctrl <- if (n_ctrl > 0) mean(ctrl_vals) else NA_real_

      if (is_control_treat || n_treat < 2 || n_ctrl < 2) {
        tibble(
          mean_treat = mean_treat,
          n_treat    = n_treat,
          control_name = if (!is_control_treat) {
            paste0("Vehicle[", vc_name, "]")
          } else {
            "Negative control"
          },
          !!paste0("mean_", resp_name, "_ctrl") := mean_ctrl,
          n_ctrl     = n_ctrl,
          t_stat     = NA_real_,
          p_value    = NA_real_
        )
      } else {
        tt <- try(stats::t.test(x, ctrl_vals, var.equal = FALSE), silent = TRUE)
        if (inherits(tt, "try-error")) {
          tibble(
            mean_treat = mean_treat,
            n_treat    = n_treat,
            control_name = paste0("Vehicle[", vc_name, "]"),
            !!paste0("mean_", resp_name, "_ctrl") := mean_ctrl,
            n_ctrl     = n_ctrl,
            t_stat     = NA_real_,
            p_value    = NA_real_
          )
        } else {
          tibble(
            mean_treat = mean_treat,
            n_treat    = n_treat,
            control_name = paste0("Vehicle[", vc_name, "]"),
            !!paste0("mean_", resp_name, "_ctrl") := mean_ctrl,
            n_ctrl     = n_ctrl,
            t_stat     = unname(tt$statistic),
            p_value    = tt$p.value
          )
        }
      }
    }) %>%
    ungroup() %>%
    rename(!!paste0("mean_", resp_name, "_treat") := mean_treat) %>%
    arrange(treatment) %>% 
  
    dplyr::filter(control_name != "Negative control")

  out
}

```


```{r}
# Functions that we use to reorganize the score table so that it follows an ascending order
library(dplyr)
reorder_wide_columns <- function(df) {
  fixed_cols <- "Compound"
  data_cols <- setdiff(names(df), fixed_cols)
  
  sort_values <- sapply(data_cols, function(x) {
    val_str <- sub(".*_([0-9.]+)X$", "\\1", x)
    val_num <- as.numeric(val_str)
    
    return(ifelse(is.na(val_num), 0, val_num))
  })
  
  sorted_data_cols <- data_cols[order(sort_values)]
  
  df %>%
    dplyr::select(all_of(fixed_cols), all_of(sorted_data_cols))
}
```

```{r}
alb_ratio_overall_14 <- ratio_overall_by_treatment_ttest(data_alb_flagged_14, "ALB")
alb_wide_T_14  <- make_level_wide(alb_ratio_overall_14,  "t_stat", "ALB_DAY14")

#If we need the reorganize the table by concentration level:
#alb_wide_T_14 <-reorder_wide_columns(alb_wide_T_14)
```

```{r}
alt_ratio_overall_T <- ratio_overall_by_treatment_ttest(data_alt_flagged, "ALT")
alt_wide_T  <- make_level_wide(alt_ratio_overall_T,  "t_stat", "ALT")
```

```{r}
atp_ratio_overall_T <- ratio_overall_by_treatment_ttest(data_atp_flagged, "ATP")
atp_wide_T  <- make_level_wide(atp_ratio_overall_T,  "t_stat", "ATP")
atp_wide_T
```
```{r}
#Combined seperate tables into one table
combined_wide_T <- alb_wide_T_14 %>%
  full_join(alt_wide_T, by = "Compound")%>%
  full_join(atp_wide_T, by = "Compound") 
# %>% full_join(XXX, by = "Compound") 
# (Use this line to add any additional table you'd like to combine)



combined_wide_T



# library(writexl)
# write_xlsx(combined_wide_T, "combined_wide.xlsx")
```
```{r}
# Function for generating heatmap
plot_tox_heatmap_all <- function(df,
                                 title    = "Treatment-to-Control",
                                 subtitle = "Sum of Ratios (relative to Vehicle)") {

  heat_df <- df %>%
    dplyr::rename(RowKey = Compound) %>%
    tidyr::pivot_longer(
      cols = -RowKey,
      names_to  = "AssayLabel",
      values_to = "Value"
    ) %>%
    dplyr::mutate(
      AssayLabel = stringr::str_replace_all(AssayLabel, "\\s+", " "),
      AssayLabel = stringr::str_replace_all(AssayLabel, "\\( ", "("),
      AssayLabel = stringr::str_replace_all(AssayLabel, " \\)", ")")
    )

  desired_rows <- c("compound 3", "compound 2", "compound 1")
  row_levels <- c(
    desired_rows[desired_rows %in% heat_df$RowKey],
    setdiff(unique(heat_df$RowKey), desired_rows)
  )

  all_assays <- unique(heat_df$AssayLabel)

  plot_df <- heat_df %>%
    dplyr::mutate(
      RowKey     = factor(RowKey, levels = row_levels),
      AssayLabel = factor(AssayLabel, levels = all_assays)
    )

  limits <- c(0, 30) #Change the min and the max of the colorbar


  
  p <- ggplot2::ggplot(plot_df, ggplot2::aes(x = AssayLabel, y = RowKey, fill = Value)) +
    ggplot2::geom_tile(color = "#1e1e1e", linewidth = 0.2) +
    ggplot2::geom_text(
      ggplot2::aes(label = sprintf("%.2f", Value)),
      size = 3, colour = "black"
    ) +
    ggplot2::scale_fill_gradientn(
      colours = c("#91bfdb", "white", "#d73027"),     
      # Change the color on the heatmap
      values  = scales::rescale(c(0, 5, 30)),         
      # Define custom breakpoints so that colors change at specific values 
      limits  = limits,
      oob     = scales::squish,
      name    = "T-score"
    ) +
    ggplot2::coord_fixed(ratio = 0.7) +
    ggplot2::labs(
      x = NULL,
      y = NULL,
      title    = title,
      subtitle = subtitle
    ) +
    ggplot2::theme_minimal(base_size = 12) +
    ggplot2::theme(
      panel.grid   = ggplot2::element_blank(),
      axis.text.x  = ggplot2::element_text(angle = 45, vjust = 1, hjust = 1, size = 11),
      axis.text.y  = ggplot2::element_text(size = 9),
      plot.title   = ggplot2::element_text(face = "bold", size = 18, colour = "#6b46c1"),
      legend.title = ggplot2::element_text(face = "bold"),
      legend.text  = ggplot2::element_text(size = 10),
      plot.margin  = ggplot2::margin(10, 16, 10, 10)
    )

  return(p)
}


```

```{r}
# Load data for heatmap
heatmap <- read_excel("combined_wide.xlsx")

# generate the heatmap
p<-plot_tox_heatmap_all(heatmap,
                 title = "Compound Toxicity Summary",
                 subtitle = "Sum of T-Scores (relative to Vehicle)")

# Save the heatmap into JPG file
ggsave(
  filename = "Compound_Toxicity_Heatmap.jpg",
  plot     = p,
  width    = 10,   
  height   = 8,     
  dpi      = 300    
)

```
